---
title: "Multiple Imputation"
output:
  html_document: default
  html_notebook: default
---

```{r setup, include=FALSE, error=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, error = TRUE)
```

Source: http://thomasleeper.com/Rcourse/Tutorials/mi.html

```{r}
n <- 10
bad <- 3
good <- n - bad

set.seed(10)
x <- c(sample(1:n, good, replace = TRUE), rep(NA, bad))
x
```

```{r}
mean(x, na.rm = TRUE)
```

Standard Error of only the elements that are not NA:
```{r}
sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
```
Generate 15 samples of new data replacing the NA elements with values extracted from the GOOD values.



```{r}
na.count <- length(x[is.na(x) == TRUE])    # calculate the # of NAs

imp <- replicate(15, 
                 c(x[!is.na(x)],                           # get the 7 non-NA elements
                   sample(x[!is.na(x)], na.count, TRUE)),  # get 3 elements from non-NA vec
                 simplify = FALSE)                         # output to be a vector
imp
```


## Applying the concept to the `activity` dataset
The dataset has to be transformed and put the intervals as variables and the steps as values. The date is the ID or index.

```{r}
library(reshape)
library(RepDataPeerAssessment1)

data(activity)

```

### Renaming `steps` as `values`
```{r}
#activity$interval <- as.factor(activity$interval)
names(activity) <- c("value", "date", "interval")
actbycol <- cast(activity, date ~ interval)
```

### Keeping original variable names
```{r}
actbycol <- cast(activity, date ~ interval, value = "steps")
```


```{r results='hold'}
cat("# of variables:", (ncol(actbycol)), "\t")
cat("# of intervals:", (ncol(actbycol)-1))
cat("\n")
mysample <- actbycol[sample(1:nrow(actbycol), 10, replace=FALSE), c(1:7, 283:289)]
mysample
```

```{r}
splice <- list(rows = c(1:5, 57:61), cols = c(1:5, 286:289))
mysample <- actbycol[splice$rows, splice$cols]
mysample
```


### Starting with interval `0` of activity by intervals
In our case, the vector to investigate is any of the interval columns.
Let's start with interval 0 assigning it as a vector.

```{r}
x0 <- actbycol$`0`
x0
```

The mean is:
```{r}
mean(x0, na.rm = TRUE)
```
The standard error:
```{r}
sd(x0, na.rm = TRUE) / sqrt(sum(!is.na(x0)))
```

```{r}
x <- x0
na.count <- length(x[is.na(x) == TRUE])    # calculate the # of NAs

imp <- replicate(15, 
                 c(x[!is.na(x)],                           # get the 7 non-NA elements
                   sample(x[!is.na(x)], na.count, TRUE)),  # get 3 elements from non-NA vec
                 simplify = FALSE)                         # output to be a vector
imp
```

  
```{r}
calcAfter <- function(x, samples = 15) {
  na.count <- length(x[is.na(x) == TRUE])    # calculate the # of NAs

  imp <- replicate(samples, c(x[!is.na(x)],
                     sample(x[!is.na(x)], na.count, TRUE)),  
                   simplify = FALSE)                  
  
  means <- sapply(imp, mean)
  grandm <- mean(means)
  
  ses <- sapply(imp, sd)/sqrt(10)
  within <- mean(ses)
  between <- sum((means - grandm)^2)/(length(imp) - 1)
  
  grandvar <- within + ((1 + (1/length(imp))) * between)
  grandse <- sqrt(grandvar)
  
  list(grandMeans = grandm, grandSE = grandse)
}
```


```{r}
calcAfter(x0, 15)
```


Now, let's see if something changed:

```{r}
means <- sapply(imp, mean)
means
```

```{r}
grandm <- mean(means)
grandm
```

```{r}
ses <- sapply(imp, sd)/sqrt(10)
within <- mean(ses)
between <- sum((means - grandm)^2)/(length(imp) - 1)

grandvar <- within + ((1 + (1/length(imp))) * between)
grandse <- sqrt(grandvar)
grandse
```

### Applying on interval `5` but with a function

```{r}
calcBefore <- function(x) {
  na.count <- length(x[is.na(x) == TRUE])    # calculate the # of NAs
  notna.count <- length(x[is.na(x) == FALSE]) 
  mean <- mean(x, na.rm = TRUE)
  serror <- sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
  list(notna = notna.count, na = na.count, mean = mean, stdError = serror)
}
```

```{r}
calcBefore(x0)
```

```{r}
x5 <- actbycol$`5`
calcBefore(x5)
calcAfter(x5)
```

```{r}
x10 <- actbycol$`10`
calcBefore(x10)
calcAfter(x10)
```


```{r}
x25 <- actbycol$`25`
calcBefore(x25)
calcAfter(x25)
```

## Vectorizing functions

### `lapply` example
```{r}
y <- list(a = 1:5, b = rnorm(10))
y
lapply(y, mean)
```
### `sapply` example

```{r}
y <- list(a = 1:5, b = rnorm(10))
y
sapply(y, mean)
```


```{r}
# subset of only the numerical data, no date
actbycol2 <- actbycol[, 2:289]
```


Applying a function to all the dataframe, excluding the date""
```{r}
sapply(actbycol2, calcBefore)
```

### Calculating the Standard Error

```{r}
sapply(actbycol2, calcAfter)
```



```{r}
hist(activity$value)
```

### histogram of activity without zeros
```{r}
activity.nonzero <- subset(activity, activity$value > 0)
hist(activity.nonzero$value)
```

### Histogram of activity restored from cast() 
```{r}
actbycol.melt <- melt(actbycol, id = "date")
rownames(actbycol.melt) <- NULL
hist(subset(actbycol.melt, actbycol.melt$value > 0)$value)
```


## Function to imputting NA with random values
```{r}
random.imp <- function (a){
  missing <- is.na(a)        # logical vector. TRUE for NAs
  n.missing <- sum(missing)  # number of missing values
  a.obs <- a[!missing]       # vector of GOOD values
  imputed <- a               # copy of original vector
  
  # Take a sample from GOOD part of the vector and replace missing elements
  imputed[missing] <- sample (a.obs, n.missing, replace=TRUE)
  return (imputed)           # return complete vector
}
```

```{r}
x
indNA <- which(is.na(x))   # which rows are NA
indNA
```

```{r}
missing <- is.na(x)
missing
```

```{r}
n.missing <- sum(missing)
n.missing
```

```{r}
x.obs <- x[!missing]
x.obs
```

```{r}
imputed <- x
imputed
```

```{r}
imputed[missing] <- sample (x.obs, n.missing, replace=TRUE)
imputed
```


### Running function on a vector

```{r}
imp <- random.imp(x)
plot(density(imp[indNA]), main = "Observed and imputed values", xlab = "x")
lines(density(imp[which(!((1:nrow(array(x))) %in% indNA))]), col = "red")
legend("topleft",text.col = c("black","red"),legend = c(" Imputation","Observed values"))
```

### Running function on 1st interval of `actbycol2`

```{r}
x0
imp <- random.imp(x0)      # impute random values
indNA <- which(is.na(x0))   # which rows are NA

plot(density(imp[indNA]), 
     main = "Observed and imputed values", 
     xlab = "x0")
lines(density(imp[which(!((1:nrow(array(x0))) %in% indNA))]), col = "red")
legend("topright",text.col = c("black","red"),legend = c(" Imputation","Observed values"))
```

```{r}
actbycol2[splice$rows, splice$cols-1]
```

### replaced by random values
```{r}
actbycol.rep <- random.imp(actbycol2[, 1:10])
```

```{r}
actbycol.rep[splice$rows, 1:10]
```

### numerical dataframe with replaced random values

```{r}
summary(actbycol2)
actbycol.imp <- random.imp(actbycol2)    
actbycol.imp[splice$rows, splice$cols-1]
```



```{r}
date <- actbycol[, 1]
actbycol.bind <- cbind(date, actbycol.imp)
```



```{r}
actbycol.imp.melt <- melt(actbycol.bind, id = "date")
names(actbycol.imp.melt) <- c("date", "interval", "steps")
# rownames(actbycol.melt) <- NULL
hist(subset(actbycol.imp.melt, actbycol.imp.melt$steps > 0)$steps)
#hist(actbycol.imp.melt$steps)
```

```{r}
dim(activity)
dim(actbycol.imp.melt)
```

## Imputing random values. Density plot

```{r}
library(reshape)
library(lattice)
library(mice)

# load the data
data("activity")

# convert interval variable to multiple columns
actbycol <- cast(activity, date ~ interval, value = "steps")

# imputting random values
actbycol.imp <- random.imp(actbycol[2:289]) 

# rebuild to interval to one column
date <- actbycol[, 1]                       # get the date
actbycol.bind <- cbind(date, actbycol.imp)  # merge 288 cols with date
actbycol.imp.melt <- melt(actbycol.bind, id = "date")  # go back to long df
names(actbycol.imp.melt) <- c("date", "interval", "steps")  # rename vars

# order by date so it is the same as original
actbycol.imp.melt <- actbycol.imp.melt[order(actbycol.imp.melt$date), ]

# convert interval to numeric
actbycol.imp.melt$interval <- as.numeric(as.character(actbycol.imp.melt$interval))

# prepare for plotting
imp.df <- actbycol.imp.melt                 # the imputted data frame
indVar <- "steps"                           # variable for plotting
indNA <- which(is.na(activity[, indVar]))   # which rows are NA
indNA.not <- which(!((1:nrow(activity)) %in% indNA))  # 
                  
plot(density(imp.df[indNA, indVar]), 
     main = "Observed and imputed values", 
     xlab = "steps")
lines(density(imp.df[indNA.not, indVar]), 
      col = "red")
legend("topright",text.col = c("black","red"),legend = c(" Imputation","Observed values"))

```

```{r}

lwd <- 1.5
data <- imp.df[, c("steps", "interval")]
tp <- xyplot(steps~interval, data = imp.df,
             na.groups = ici(imp.df),
             ylab ="steps", xlab="interval",
       cex = 0.75, 
       xlim = c(0, 2355),
       
       lex=lwd
       #ylim = c(0, 900), 
       #xlim = c(0, 500))
)
#print(tp, newpage = FALSE, position = c(0.48,0.08,1,0.92))
tp
```

```{r}

```


### Summary
```{r}
info.ints <- data.frame(rows = nrow(imp.df), 
                   cols = ncol(imp.df),
                   min = min(imp.df[, indVar]),
                   max = max(imp.df[, indVar]),
                   NAs.before = length(indNA),
                   NAs.after = length(indNA.not)
                   )
          
info.dbls <- data.frame(median = median(imp.df[, indVar]),
                        mean = mean(imp.df[, indVar]),
                        sd = sd(imp.df[, indVar]),
                        ses = sd(imp.df[, indVar]) / 
                                  sqrt(sum(!is.na(imp.df[, indVar])))
                   )
info.ints <- t(info.ints)
info.dbls <- t(info.dbls)
info.ints
round(info.dbls, 3)
```

### Normalized variables `steps` and  `row-index`
```{r}
y <- imp.df$steps
y.sig <- sd(y)
y.mu <- mean(y)
y.norm <- (y - y.mu) / y.sig
min(y.norm)
max(y.norm)

x <- seq(1:nrow(imp.df))
x.sig <- sd(x)
x.mu <- mean(x)
x.norm <- (x - x.mu) / x.sig
min(x.norm)
max(x.norm)
```

### Normalized interval (optional) DO NOT RUN
This has the disadvantage of showing discontinuities along the x-axis since the interval is no n contiguous jumping in steps of 5.

```{r}
x <- imp$interval
x.sig <- sd(x)
x.mu <- mean(x)
x.norm <- (x - x.mu) / x.sig
min(x.norm)
max(x.norm)
```

### Normalized plot `x` vs `y` of random imputting
Normalizing the row-index should be the prefferred option to avoid showing gap along the x-axis.

```{r fig.asp=1}
# find NA and not NA row-indexes
indVar <- "steps"
indNA <- which(is.na(activity[, indVar]))   # which rows are NA
indNA.not <- which(!((1:nrow(activity)) %in% indNA))

# plot row-index vs steps
col <- c(mdc(4),mdc(5))

plot(x.norm[indNA.not], y.norm[indNA.not], 
     main = "index vs steps",
     xlab = "x",
     ylab = "y",
     col = col[1]
     )
points(x.norm[indNA], y.norm[indNA], col = col[2])
```


## Mean Imputation

```{r}
library(RepDataPeerAssessment1)
library(lattice)
library(mice)

data(activity)

activity.nodate <- activity[, c(1, 3)]

### impute the mean
#imp <- mice(activity.nodate, method="mean", m=1, maxit=1)
imp <- mice(activity.nodate, method="mean", m=1, maxit=1)

### Figure 1.1
lwd <- 1.5
# par(mfrow=c(1,2))

max <- max(nas.complete(activity$steps))

breaks <- seq(0, max*1.1, 20)
nudge <- 1

x <- matrix(c(breaks-nudge, breaks+nudge), ncol=2)

obs <- activity[, "steps"]
mis  <- imp$imp$steps[,1]

fobs <- c(hist(obs, breaks, plot=FALSE)$counts, 0)
fmis <- c(hist(mis, breaks, plot=FALSE)$counts, 0)

y <- matrix(c(fobs, fmis), ncol=2)

matplot(x, y, type="s",
        col=c(mdc(4),mdc(5)), lwd=2, lty=1,
        #xlim = c(0, 170), 
        ylim = c(-5, 200), 
        yaxs = "i",
        xlab="Steps",
        ylab="Frequency")
# box()


```

### Plot `interval` vs `steps` for mean imputting
```{r fig.asp=1}
tp <- xyplot(imp, steps~interval, 
             na.groups = ici(imp),
             ylab ="steps", xlab="interval",
       cex = 0.75, 
       
       lex=lwd
       #ylim = c(0, 900), 
       #xlim = c(0, 500))
)
#print(tp, newpage = FALSE, position = c(0.48,0.08,1,0.92))
tp
```



```{r fig.asp=1}

# find NA and not NA row-indexes
indVar <- "steps"
indNA <- which(is.na(activity[, indVar]))            # which rows are NA
indNA.not <- which(!((1:nrow(activity)) %in% indNA)) # rows that are not NA

steps.imp <- imp$imp$steps$`1`   # all imputed values ONLY; no observations
steps <- activity$steps          # obs + NAs

# summary(steps); dim(steps)
steps[indNA] <- steps.imp        # replace the NAs with imputed values
# summary(steps); dim(steps)


y <- steps
y.sig <- sd(y)
y.mu <- mean(y)
y.norm <- (y - y.mu) / y.sig
min(y.norm)
max(y.norm)

x <- seq(1:nrow(activity))
x.sig <- sd(x)
x.mu <- mean(x)
x.norm <- (x - x.mu) / x.sig
min(x.norm)
max(x.norm)



# plot row-index vs steps
col <- c(mdc(4),mdc(5))

plot(x.norm[indNA.not], y.norm[indNA.not], 
     main = "index vs steps",
     xlab = "x",
     ylab = "y",
     col = col[1]
     )
lines(x.norm[indNA], y.norm[indNA], col = col[2], lwd = 2)
points(x.norm[indNA], y.norm[indNA], col = col[2])
# abline(y = y.mu, lwd = 3)
```

## Regression Imputation
```{r}

fit <- lm(steps ~ interval, data=activity)
pred <- predict(fit, newdata=ic(activity))

### alternative using mice
imp <- mice(activity[,c(1,3)], method="norm.predict", m=1, maxit=3,seed=1)
```

```{r}
### Figure 1.2
par(mfrow = c(1,2))

fmis <- c(hist(pred, breaks, plot=FALSE)$counts, 0)
y <- matrix(c(fobs, fmis), ncol=2)

matplot(x, y, type="s",
        col=c(mdc(4),mdc(5)), lwd=2, lty=1,
        ylim = c(0, 200),
        yaxs = "i",
        xlab="steps",
        ylab="Frequency")
box()

tp <- xyplot(imp, steps~interval,
       ylab="steps", xlab="interval",
       cex = 0.75, lex=lwd,
)
print(tp, newpage = FALSE, position = c(0.48,0.08,1,0.92))
```


## Stochastic regression imputation

```{r}

library(lattice)
library(mice)

imp <- mice(activity[,c(1,3)], method="pmm", m=1, maxit=3, seed=1)

lwd <- 2
nudge <- 1
### Figure 1.3
par(mfrow = c(1,2))

mis  <- imp$imp$steps[,1]

breaks <- seq(-500, 900, 10)

obs <- activity[,"steps"]

fobs <- c(hist(obs, breaks, plot=FALSE)$counts, 0)
fmis <- c(hist(mis, breaks, plot=FALSE)$counts, 0)

x <- matrix(c(breaks-nudge, breaks+nudge), ncol=2)
y <- matrix(c(fobs, fmis), ncol=2)

matplot(x, y, type="s",
        col=c(mdc(4),mdc(5)), lwd=2, lty=1,
        ylim = c(0, 300),
        yaxs = "i",
        xlab="steps",
        ylab="Frequency")

box()

tp <- xyplot(imp, steps~interval, na.groups=ici(imp),
       ylab="steps", xlab="interval",
       cex = 0.75, lex=lwd,
       )
# tp
print(tp, newpage = FALSE, position = c(0.48,0.08,1,0.92))
```

## LOCF

```{r}
library(lattice)
library(mice)
library(zoo)

St <- activity$steps

locf <- function(x) {
  # last observation carried forward
  NonNAindex <- which(!is.na(x))  # find the index of all non-NAs
  firstNonNA <- min(NonNAindex)   # get the lower index or 1st non-NA
  a <- x[firstNonNA]              # assign 1st non-NA observation to pivot
  
  for (i in 1:length(x)) {        # count from the 1st to the end of the vector
    if (is.na(x[i])) x[i] <- a    # if current element is NA, then make it as pivot
    else a <- x[i]                # else make pivot same as ith
  }
  return(x)
}

Sti <- locf(St)
length(Sti)
nas.count(Sti)
colvec <- ifelse(is.na(St), mdc(2), mdc(1))

### Figure 1.4

limRow <- 12000

plot(Sti[1:limRow], 
     col = colvec, 
     type="l", 
     xlab="Observation", 
     ylab="Steps")

points(Sti[1:limRow], col = colvec, pch=20, cex=1)
```

```{r}
LOCF <- function(x) {
    # Last Observation Carried Forward (for a left to right series)
    LOCF <- max(which(!is.na(x))) # the location of the Last Observation to Carry Forward
    x[LOCF:length(x)] <- x[LOCF]
    return(x)
}

Sti <- LOCF(St)
length(Sti)
nas.count(Sti)

par(mfrow=c(1,1))
limRow <- 17568

colvec <- ifelse(is.na(St), mdc(2), mdc(1))    # missing data coloring (Mdc)

plot(Sti[1:limRow], 
     col = colvec, 
     type="l", 
     xlab="Observation", 
     ylab="Steps")

points(Sti[1:limRow], col = colvec, pch=20, cex=1)
```


```{r}
na.locf <- function(x) {
  v <- !is.na(x)
  c(NA, x[v])[cumsum(v)+1]
}

Sti <- na.locf(St)
length(Sti)

nas.count(Sti)
par(mfrow=c(1,1))
limRow <- 12000

colvec <- ifelse(is.na(St), mdc(2), mdc(1))    # missing data coloring (Mdc)

plot(Sti[1:limRow], 
     col = colvec, 
     type="l", 
     xlab="Observation", 
     ylab="Steps")

points(Sti[1:limRow], col = colvec, pch=20, cex=1)

```


```{r}
### Figure 1.4

library(zoo)

St <- activity$steps

# Sti <- na.locf(St)
bz <- zoo(St)
length(bz)
Sti <- zoo::na.locf(bz)
length(Sti)

nas.count(Sti)
par(mfrow=c(1,1))
limRow <- length(Sti)

colvec <- ifelse(is.na(St), mdc(2), mdc(1))    # missing data coloring (Mdc)

plot(Sti[1:limRow], 
     col = colvec, 
     type="l", 
     xlab="Observation", 
     ylab="Steps")

points(Sti[1:limRow], col = colvec, pch=20, cex=1)
```

```{r}
mean(St, na.rm = TRUE)
```

```{r}
# Standard Error of only the elements that are not NA
sum(!is.na(St))
sd(St, na.rm = TRUE) / sqrt(sum(!is.na(St)))
```

```{r}
mean(Sti, na.rm = TRUE)
```

```{r}
# Standard Error of only the elements that are not NA
sum(!is.na(Sti))
sd(Sti, na.rm = TRUE) / sqrt(sum(!is.na(Sti)))
```

## Applying LOCF to a transposed data.frame

```{r}
library(reshape)
library(RepDataPeerAssessment1)

data(activity)

#activity$interval <- as.factor(activity$interval)
names(activity) <- c("value", "date", "interval")
actbycol <- cast(activity, date ~ interval)

actbycol.nodate <- actbycol[, 2:289]    # remove the date column for now

```


```{r}
act <- sapply(actbycol.nodate, zoo::na.locf)    # apply LOCF to numerical intervals
df <- data.frame(act)       # create data frame
names(df) <- colnames(act)  # rename

# rebuild to old activity structure
actbycol.shifted <- actbycol[2:61, 1]       # get only the date column but not row #1
actbycol.new <- cbind(actbycol.shifted, df)

#rownames(actbycol.new) <- NULL
```

```{r}
names(actbycol.new)[1] <- "date"
```


```{r}
activity.new <- melt(actbycol.new, id = "date")    # melt DF as original
names(activity.new) <- c("date", "interval", "steps")  # rename vars

```

```{r}
mean(activity.new$steps)
```

```{r}
sd(activity.new$steps) / sqrt(sum(!is.na(activity.new$steps)))
```
This number is not better than the one we obtained by doing the LOFC replacement in the whole steps column above.





