---
title: "Reproducible Research, Assignment 1"
author: "Alfonso R. Reyes"
date: "`r Sys.Date()`"
# output: 
#   html_document:
#     keep_md: true
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    keep_md: true

vignette: >
  %\VignetteIndexEntry{Reproducible Research, Assignment 1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Loading and preprocessing the data

```{r}

```


## What is mean total number of steps taken per day?



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?




## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r,  fig.cap="my caption"}
plot(1:10)

```


## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
