---
title: "dealing with missing data"
output: html_notebook
---

## Using own package

```{r}
library(RepDataPeerAssessment1)

data("activity")
head(activity)
```


```{r}
steps <- activity$steps
summary(steps)
```

We have 2304 NA's.

How important is it?
```{r}
mean(is.na(steps))
```

We have 13% of missing values.

```{r}
missing.days <- 
```


## Using `dplyr`

How many values are zeros?
```{r message=FALSE}
library(dplyr)

filter(activity) %>%
  summarize(zeros = mean(steps != 0, na.rm = TRUE))
```

## Where our data lies
Source: http://thomasleeper.com/Rcourse/Tutorials/NAhandling.html

Analyze a vector or variable of the dataframe:
```{r}
length(steps[is.na(steps) == TRUE])     # number of NAs
```

Analyze for NAs in a the whole dataframe:
```{r}
head(is.na(activity))
length(steps[is.na(steps) == FALSE])     # number of non-NAs
```

### Visualizing missing data in a plot
```{r, fig.asp=1}
image(is.na(activity),
      xaxt = "n",         # suppress plotting of x axis
      yaxt = "n",         # suppress plotting of y axis
      bty = "n"           # do not draw a box around
      )

# custom labels for x and y axis
axis(1, seq(0, 1, length.out = nrow(activity)), 1:nrow(activity), col = "white")
axis(2,                 # y axis
     c(0, 0.5, 1),      # sequence of labels
     names(activity),   # variables names of data frame
     col = "white",     # color
     las = 2)           # 2: perpendicular to the axis. Defaults = 0

box(lty = '1373', col = 'black')    # draw a box around
```







## Using the mean to replace NA values
Source: https://www.r-bloggers.com/example-2014-5-simple-mean-imputation/

```{r}
df = data.frame(x = 1:20, y = c(1:10,rep(NA,10)))
head(df)
tail(df)
```

```{r}
df$y[is.na(df$y)] = mean(df$y, na.rm = TRUE)
```

```{r}
tail(df)
```

### Alternative
```{r}
# alternative
df = data.frame(x = 1:20, y = c(1:10,rep(NA,10)))
head(df)
tail(df)

df = transform(df, y = ifelse(is.na(y), mean(y, na.rm=TRUE), y))
tail(df)
```

In the first example, we identify elements of y that are NA, and replace them with the mean, if so. In the second, we test each element of y; if it is NA, we replace with the mean, otherwise we replace with the original value. 



