---
title: "Reproducible Research, Assignment 1"
author: "Alfonso R. Reyes"
date: "`r Sys.Date()`"
output:
  html_document: default
  html_notebook: default
---


```{r setup, include=FALSE, error=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, error = TRUE)
```


## A. Loading and preprocessing the data
We start by downloading the raw data using the link provided by the instructor. We use a function that will download the zip file, unpack it and place it in an indicated directory. The function is called `downloadZip`.

### 1. Load the data

#### Downloading and unpacking raw data file

```{r}
library(RepDataPeerAssessment1)
library(rprojroot)

project.data <- find_package_root_file('data')
project.extdata <- find_package_root_file('inst/extdata')
project.R <- find_package_root_file('R')
project.data
project.R
project.extdata
```


```{r}
downloadZip <- function(fileUrl, outDir="./data") {
  temp <- tempfile()
  download.file(fileUrl, temp, mode = "wb")
  unzip(temp, exdir = outDir)
}
```


```{r}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
outDir <- project.extdata             # folder for raw data
downloadZip(fileUrl, outDir = outDir)   # download and unpack zip file
```


#### Create the RData file
More RData files may be generated during this assignment. They will be placed under the folder `data`.



#### Saving the dataset to `data`
Reads the CSV raw data file from `inst/extdata` and save it under `data`. Then remove it from memory.
```{r}
# save the dataset
activity.raw <- read.csv(paste(project.extdata, "activity.csv", sep = "/"))
save(activity.raw, file=paste(project.data, "activity.raw.RData", sep = "/"))
activity.raw
```

** Convert variable `date` as.factor to `as.date` **
```{r}
activity <- activity.raw
activity$date <- as.Date(activity.raw$date)
str(activity) 
```



```{r}
# save the dataset
save(activity, file=paste(project.data, "activity.RData", sep = "/"))
activity
```

```{r}
# rm(activity)
load(paste(project.data, "activity.RData", sep = "/"))
activity
```



Confirm it has been saved:
```{r}
file.exists(paste(project.data, "activity.RData", sep = "/"))
```

### 2. Process and transform
Process/transform the data (if necessary) into a format suitable for the analysis.

#### Basic sanity check

* Show first 6 rows of data frame
* Show dimensions of the data frame
* Show variable names
* Show summary

```{r message=FALSE}
library(RepDataPeerAssessment1)
data("activity")
activity
```


```{r}
activity.env <- environment()
load(paste(project.data, "activity.rda", sep = "/"), 
     envir = activity.env)
activity.fit1 <- activity.env$activity
head(activity.fit1)
```




```{r}
data(intervals)
inter <- intervals
inter
```


**<u>Show dimensions</u>**
```{r}
dim(activity)
```
**<u>Names of the variables</u>**

```{r}
names(activity)
```

**<u>Structure</u>**

```{r}
str(activity)
```

**<u>Summary</u>**
```{r}
summary(activity)
```

```{r}

```




```{r}
save(activity, file = paste(project.data, "activity.RData", sep = "/"))
rm(activity)      # remove from memory
load(paste(project.data, "activity.RData", sep = "/"))
# data(activity)    # reload 
activity
str(activity)
```
```{r}
file.remove(paste(project.data, "activity.RData", sep = "/"))
```



```{r}

```


```{r}
activity
```


## B. What is mean total number of steps taken per day?
We will ignore the NAs in this part of the assignment. 

### 1. Calculate the total number of steps taken per day

```{r}
# get only observations that are not NA
complete <- complete.cases(activity)
activity.cases <- activity[complete, ]
dim(activity.cases)
```

```{r fig.asp=1}
plot(seq(1:nrow(activity.cases)), activity.cases$steps)
```


#### Histogram of total number of steps each day

```{r}
hist(activity.cases$steps, breaks = 30)
```


#### Find the **mean** and the **median** total number of steps per day

```{r}
mean(activity.cases$steps)
```

```{r}
median(activity.cases$steps)
```

#### How many unique intervals are there?
We want to know how many unique intervals there are because we will need later to calculate the maximum steps per interval and we need this number to verify our count of intervals is correct.

```{r}
unique(activity.cases$interval)
```
So, there are `r length(unique(activity.cases$interval))` **unique** intervals, with the first interval being `r min(activity.cases$interval)` and the last `r max(activity.cases$interval)`.

-----

## C. What is the average daily activity pattern?

### 1. Make a time series plot (i.e. type = "l") 
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
byDate.steps.mean <- aggregate(activity.cases$steps, 
                               by = list(activity.cases$date), mean)

# rename the variable to something meaningful
names(byDate.steps.mean) <- c("Day", "mean.steps")
plot(byDate.steps.mean$Day, byDate.steps.mean$mean.steps, type = "l")
```

### 2. Which interval contain the maximum # of steps
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

#### Find day and interval with maximum number of steps

Let's find the maximum number of steps:
```{r}
max.steps <- max(activity.cases$steps)
max.steps
```

Now, let's find which interval holds the maximum number of steps:
```{r}
index <- which(activity.cases$steps == max.steps)
whole_row <- activity.cases[index, ]
whole_row
```


**<u>Grouping by interval</u>**
```{r}
byInterval <- aggregate(activity.cases$steps, 
                        by = list(activity.cases$interval), 
                        max)

names(byInterval) <- c("interval", "steps.max")
sorted <- byInterval[order(-byInterval$steps.max), ]   # order by steps, descending
head(sorted)
```

#### Find what are the intervals with more activity
We will plot the number maximum of steps in logarithmic scale:

```{r}
plot(sorted$interval, sorted$steps.max,
     xlim = c(0, 2500), ylim = range(1:1000),
     panel.first = grid(lty = 1))
```

```{r}
plot(sorted$interval, sorted$steps.max, log = "y",
     xlim = c(0, 2500), ylim = range(1:1000),
     panel.first = grid(lty = 1))
```

And zooming in in a non-logarithmic plot, we can see that the maximum values are those around 800.
```{r}
# non-logarithmic plot
plot(sorted$interval, sorted$steps.max,
     xlim = range(0:2500), ylim = range(700:825),
     panel.first = grid(lty = 1))
```

What we can confirm is that the maximum activity occurs between the **500 and 2000** intervals.


## D. Imputing missing values

### 1. Calculate the number of missing values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
# get only observations that are  NA
complete <- complete.cases(activity)
activity.missing <- activity[!complete, ]
nrow(activity.missing)
```

### 2. Complete missing values
Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

There are several ways we can complete the missing values. 






### 3. Create dataset with missing values filled in
Create a new dataset that is equal to the original dataset but with the missing data filled in.

### 4. Make a histogram (with missing data filled in)
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?





## E. Are there differences in activity patterns between weekdays and weekends?

### 1. Create a new factor variable
Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.


### 2. Make a panel
Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.



