---
title: "Reproducible Research, Assignment 1"
author: "Alfonso R. Reyes"
date: "`r Sys.Date()`"
output:
  html_document: default
  html_notebook: default
---


```{r setup, include=FALSE, error=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, error = TRUE)
```


## Loading and preprocessing the data
We start by downloading the raw data using the link provided by the instructor. We use a function that will download the zip file, unpack it and place it in an indicated directory. The function is called `downloadZip`.


### Downloading and unpacking raw data file

```{r}
downloadZip <- function(fileUrl, outDir="./data") {
  temp <- tempfile()
  download.file(fileUrl, temp, mode = "wb")
  unzip(temp, exdir = outDir)
}
```


```{r}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
outDir <- "../inst/extdata"             # folder for raw data
downloadZip(fileUrl, outDir = outDir)   # download and unpack zip file
```


### Create the RData file
More RData file may be generated during this assignment. They will be placed under the folder `data`.

```{r}
# get the base folder and raw data folder
baseLoc <- system.file(package="RepDataPeerAssessment1")
extPath <- file.path(baseLoc, "extdata")
baseLoc
extPath
```

```{r}
# save the 
activity <- read.csv(paste(extPath, "activity.csv", sep = "/"))
save(activity, file="../data/activity.RData")
rm(activity)
```


```{r}
# this will work after at least a data file has been saved to this folder
system.file(file="data", package="RepDataPeerAssessment1")
```

### Basic sanity check

* Show first 6 rows of data frame
* Show dimensions of the data frame
* Show variable names
* Show summary

```{r message=FALSE}
library(RepDataPeerAssessment1)
data("activity")
head(activity)
```


```{r}
dim(activity)
```

```{r}
names(activity)
```

```{r}
str(activity)
```

```{r}
summary(activity)
```

### Convert variable `date` as.factor to as.date

```{r}
activity$date <- as.Date(activity$date)
save(activity, file="../data/activity.RData")
str(activity)
```


## What is mean total number of steps taken per day?
We will ignore the NAs in this part of the assignment. 

```{r}
# get only observations that are not NA
complete <- complete.cases(activity)
activity.cases <- activity[complete, ]
```


### Histogram of total number of steps each day

```{r}
hist(activity.cases$steps, breaks = 60)
```

```{r}
hist(activity.cases$steps, breaks = 30)
```


### Find the **mean** and the **median** total number of steps per day

```{r}
mean(activity.cases$steps)
```

```{r}
median(activity.cases$steps)
```

### How many unique intervals are there?
We want to know how many unique intervals there are because we will need later to calculate the maximum steps per interval and we need this number to verify our count of intervals is correct.

```{r}
unique(activity.cases$interval)
```
So, there are `r length(unique(activity.cases$interval))` intervals.



## What is the average daily activity pattern?

### Time series plot (day vs avg. steps)
(i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis) 

```{r}
plot(activity.cases$interval, activity.cases$steps, type = "l")
```





```{r}
byDate.steps.mean <- aggregate(activity.cases$steps, by = list(activity.cases$date), mean)
# rename the variable to something meaningful
names(byDate.steps.mean) <- c("Day", "mean.steps")
plot(byDate.steps.mean$Day, byDate.steps.mean$mean.steps, type = "l")
```

### Find day and interval with maximum number of steps
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


```{r}
max(activity.cases$steps)
```



```{r}
max.steps <- max(activity.cases$steps)
max.steps
```


```{r}
index <- which(activity.cases$steps == max.steps)
whole_row <- activity.cases[index, ]
whole_row
```



```{r}
# activity.cases$interval[which(activity.cases$steps == max(activity.cases$steps))]
```

```{r}
# Find max steps per day and plot
byDate.steps.max <- aggregate(activity.cases$steps, by = list(activity.cases$date), max)
names(byDate.steps.max) <- c("Day", "steps.max")
plot(byDate.steps.max$Day, byDate.steps.max$steps.max, type = "l")
```

By interval

```{r}
byInterval <- aggregate(activity.cases$steps, 
                        by = list(activity.cases$interval), 
                        max)

sorted <- byInterval[order(-byInterval$x), ]
```

```{r}
names(byInterval)
```


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?





## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
