---
title: "imputeTestbench pkg"
output: html_notebook
---

Source: https://github.com/neerajdhanraj/imputeTestbench/blob/master/Manuscript%20imputeTestbench.pdf

The default dataset if dataIn = NULL is nottem, a time series object of average
air temperatures recorded at Nottingham Castle from 1920 to 1930. This dataset is included with the base datasets package.

```{r}
library(imputeTestbench)
set.seed(123)

a <- impute_errors()
a
```

The errprof object is a list with seven elements. The ﬁrst element, Parameter, is a character string
of the error metric used for comparing imputation methods. The second element, MissingPercent, is
a numeric vector of the missing percentages that were evaluated in the input dataset. The remaining
ﬁve elements show the average error for each imputation method at each interval of missing data
in MissingPercent. The averages at each interval are based on the repetitions speciﬁed in the initial
call to impute_errors() where the default is Repetition = 1 . Although the print method for the
errprof object returns a list, the object stores the unique error estimates for every imputation method,
repetition, and missing data interval. These values are used to estimate the averages in the printed
output and to plot the distribution of errors with plot_errors() shown below. 

```{r}
plot_errors(a)
```

```{r}
plot_errors(a, plotType = 'line')
```


```{r}
plot_impute(showmiss = T)
```

## Demonstration of imputeTestbench with examples
The austres dataset is a ts object of Australian population in thousands, measured quarterly
from 1971 to 1994 (Brockwell and Davis, 1996). The plot_errors() function shows that all imputation
methods had larger error values with additional missing observations, as expected, and that the
na.mean imputation method had the largest error values. Differences between the error values can
be understood by viewing a sample of the imputed data with plot_impute(). The example belows
shows an example of imputed values using the na.approx, na.locf, and na.mean functions at 10%
and 90% missing observations using MCAR sampling.

```{r}
aus <- datasets::austres
ex <- impute_errors((dataIn = aus))
ex
```

```{r}
plot_errors(ex, plotType = 'line')
```

```{r}
plot_impute(methods = c("na.mean", "na.locf", "na.approx"), missPercent = 90)
```

## Sample
Source: http://www.neerajbokde.com/cran/imputetestbench

```{r}
datax <- c(1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5,1:5)
datax
```

```{r}
library(imputeTestbench)

q <- impute_errors(datax)
q
```

```{r}
plot_errors(q)
```

```{r}
plot_errors(dataIn = q, plotType = "line")
```

