---
title: "Untitled"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

## Loading required packages
```{r eval = TRUE}
require(ggplot2)
```

## Loading and preprocessing the data

```{r}
if(!file.exists("./data")){dir.create("./data")}
fileUrl = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl, destfile = "./data/Activity-monitoring-data.zip")

dl = unzip(zipfile = "./data/Activity-monitoring-data.zip", exdir = "./data")
dt.Raw = read.csv(dl)
```

## What is the Mean total numbers of steps taken per day?
#### For this part of the assignment we're ignoring missing values in the dataset

#### 1 - Calculating the total number of steps taken per day

```{r}
dt.Steps = aggregate(list(Steps = dt.Raw$steps), by = list(Date = dt.Raw$date), FUN = sum, na.rm = TRUE)

dt.Steps
```

### 2 - Creating a Histogram of total number of steps taken each day
```{r}
hist(dt.Steps$Steps, breaks = seq(0,25000, by = 2500), col = "lightgreen", main = "Frequency of Steps per Day", xlab = "Steps per Day (Groups of 2500)", ylab = "Frequency")
```

#### 3 - Calculating the mean & median number of steps per day

```{r}
mean(dt.Steps$Steps)
median(dt.Steps$Steps)
```

## What is the average daily activity pattern?
#### For this part of the assignment we're continuing to ignore missing values in the dataset

```{r}
dt.DailyActivityPattern = aggregate(list(Steps = dt.Raw$steps), by = list(Interval = dt.Raw$interval), FUN = mean, na.rm = TRUE)
```

#### 1 - Making a time series plot of the average steps taken in every 5-minute interval

```{r}
plot(dt.DailyActivityPattern$Interval, dt.DailyActivityPattern$Steps, type = "l", col = "lightgreen", main = "Avg. Steps per 5-minute Interval", xlab = "5-minute Intervals", ylab = "Avg. Steps Taken")
```

#### 2 - Finding which interval, on average, has the maximum number of steps taken per day

```{r}
dt.DailyActivityPattern[which.max(dt.DailyActivityPattern$Steps),]
```

## Imputing missing values

#### 1 - Calculating total number of missing values in the dataset

```{r}
sum(is.na(dt.Raw$steps))
```

#### 2/3 - Replacing all missing values with the avg. steps of that interval

```{r}
dt.NA = is.na(dt.Raw$steps)
dt.AvgInterval = tapply(dt.Raw$steps, dt.Raw$interval, mean, na.rm = TRUE, simplify = TRUE)
dt.Imputed = dt.Raw
dt.Imputed$steps[dt.NA] = dt.AvgInterval[as.character(dt.Imputed$interval[dt.NA])]
```

#### Recalculating total number of missing values in the dataset

```{r}
sum(is.na(dt.Imputed$steps))
```

#### 4a - Creating a Histogram of total number of steps taken each day

```{r}
dt.StepsImputed = aggregate(list(Steps = dt.Imputed$steps), by = list(Date = dt.Imputed$date), FUN = sum, na.rm = TRUE)

hist(dt.StepsImputed$Steps, breaks = seq(0,25000, by = 2500), col = "lightgreen", main = "Frequency of Steps per Day", xlab = "Steps per Day (Groups of 2500)", ylab = "Frequency")
```

#### 4b - Calculating the mean & median number of steps per day

```{r}
mean(dt.StepsImputed$Steps)
median(dt.StepsImputed$Steps)
```

## Are there differences in activity patterns between weekdays and weekends?

#### 1 - Creating new variable "DateType"

```{r}
dt.Dates = dt.Imputed
dt.Dates$Day = weekdays(as.POSIXlt(dt.Imputed$date))
dt.Dates$DateType = ifelse(dt.Dates$Day %in% c("Saturday","Sunday"),"Weekend", "Weekday")
```

#### 2 - Creating a panel plot with average number of steps taken per interval on Weekdays and Weekends

```{r}
dt.DatesAgg = aggregate(data = dt.Dates, steps ~ interval + DateType, mean)

ggplot(dt.DatesAgg, aes(interval, steps, color = DateType)) + 
        geom_line() + 
        facet_grid(DateType ~ .) +
        xlab("5-minute Intervals") + 
        ylab("Avg. Steps Taken")
```